---
title: "GENEActiv and Phase Angle in IIM — Bootstrap Mixed-Model Report"
author: "Veronika Balounová"
date: "Compiled on: **2026-05-04**  <br> Last updated **`r format(Sys.time(), '%Y-%m-%d')`**"
output:
  html_document:
    code_folding: hide
    df_print: paged
    theme: yeti
    highlight: tango
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
  pdf_document:
    fig_caption: yes
    toc: yes
editor_options:
  markdown:
    wrap: 72
---

```{=html}
<style type="text/css">
.title {
  text-align: center;
}
.author {
  font-size: 30px;
  text-align: center;
  font-style: italic;
  font-weight: 300;
}
.date {
  font-weight: 300;
}
h1, h2, h3, h4 {
    font-weight: 600;
    margin-top: 2em;
}
body {text-align: justify}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    border-left: 5px solid #eee;
}
p {
  text-indent: 2rem;
}
table {
    width:100%;
}
.main-container {
  max-width: 100% !important;
  margin: auto;
}
img{
    height: 100%;
    width: 100%;
    object-fit: contain;
}
.boxBorder {
     border: 2px solid #990066;
     padding: 10px;
     outline: #990066 solid 5px;
     outline-offset: 5px;
   }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  knitr.kable.NA = ""
)
options(knitr.kable.NA = "", scipen = 999)

if (!require("pacman")) install.packages("pacman")

pacman::p_load(update = FALSE,
  tidyverse, readxl, janitor, gt, knitr, kableExtra,
  lme4, parameters, performance, datawizard, sessioninfo
)

conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::mutate,
  dplyr::rename,
  dplyr::summarize,
  dplyr::summarise,
  dplyr::select,
  dplyr::lag,
  purrr::map
)

rmd_dir <- normalizePath(dirname(knitr::current_input(dir = TRUE)), winslash = "/")

path_boot <- "R:/MYOZITIDY_VÝZKUM/_IMMET-GRANT/GENEActivII/lmer_custom_multivariable_phase_angle/residualizace/bootstrap"
path_input <- "R:/MYOZITIDY_VÝZKUM/_IMMET-GRANT/GENEActivII/rCCA_rozdelene_sady_parametru/ggir_vs_myokiny_bez_imputace/20263003.xlsx"

file_xlsx <- file.path(path_boot, "bootstrap_lmer_report.xlsx")
file_summary <- file.path(path_boot, "bootstrap_summary.csv")
file_status <- file.path(path_boot, "bootstrap_status.csv")
file_all <- file.path(path_boot, "bootstrap_all_parameters.csv")
file_full_model <- file.path(path_boot, "full_sample_model_raw.rds")

fmt <- function(x, digits = 3) {
  ifelse(is.na(x), NA_character_, formatC(x, format = "f", digits = digits))
}

fmt_p <- function(x) {
  dplyr::case_when(
    is.na(x) ~ NA_character_,
    x < 0.001 ~ "<0.001",
    TRUE ~ formatC(x, format = "f", digits = 3)
  )
}

parameter_labels <- c(
  "(Intercept)" = "Intercept (baseline M0, female reference)",
  "examM3" = "M3 vs M0",
  "examM6" = "M6 vs M0",
  "examM18" = "M18 vs M0",
  "examM30" = "M30 vs M0",
  "examM48" = "M48 vs M0",
  "age" = "Age",
  "age_z" = "Age, z-score",
  "sexMale" = "Male sex vs female",
  "bcm_between" = "Between-patient BCM",
  "bcm_between_z" = "Between-patient BCM, z-score",
  "bmi_between" = "Between-patient BMI",
  "bmi_between_z" = "Between-patient BMI, z-score"
)

summary_tbl <- readr::read_csv(file_summary, show_col_types = FALSE) |>
  mutate(
    parameter_label = coalesce(parameter_labels[Parameter], Parameter),
    ci = paste0(fmt(ci_low_percentile), " to ", fmt(ci_high_percentile)),
    estimate = fmt(estimate_median),
    boot_se_fmt = fmt(boot_se),
    p_fmt = fmt_p(p_boot_two_sided),
    direction = case_when(
      ci_low_percentile > 0 ~ "positive",
      ci_high_percentile < 0 ~ "negative",
      TRUE ~ "uncertain"
    )
  )

status_tbl <- readr::read_csv(file_status, show_col_types = FALSE)

full_params <- readxl::read_excel(file_xlsx, sheet = "full_sample_parameters_raw") |>
  mutate(
    parameter_label = coalesce(parameter_labels[Parameter], Parameter),
    ci = paste0(fmt(CI_low), " to ", fmt(CI_high)),
    p_fmt = fmt_p(p)
  )

full_collin <- readxl::read_excel(file_xlsx, sheet = "full_sample_collinearity_raw")

input_data <- readxl::read_excel(path_input, sheet = "sheet1") |>
  janitor::clean_names() |>
  rename(exam = any_of("m")) |>
  mutate(
    exam = factor(exam, levels = c(0, 3, 6, 18, 30, 48),
                  labels = c("M0", "M3", "M6", "M18", "M30", "M48")),
    sex_label = factor(sex, levels = c(0, 1), labels = c("Female", "Male"))
  )

eligible_data <- input_data |>
  filter(!is.na(filename_2)) |>
  datawizard::demean(select = c("bcm", "bmi"), by = "filename_2") |>
  filter(
    !is.na(ig_gradient_enmo),
    !is.na(exam),
    !is.na(age),
    !is.na(sex),
    !is.na(bcm_between),
    !is.na(bmi_between)
  )
```

# Aim

This report summarizes bootstrap mixed-model results from the
`lmer_srovnani_2_promennych_interakce_bootstrap_bmi.R` analysis. The
clinical question is whether longitudinal wrist-accelerometry output
(`ig_gradient_enmo`) changes over time in patients with idiopathic
inflammatory myopathy (IIM), and whether the signal is independently
associated with age, sex, between-patient body cell mass (BCM), and
between-patient body mass index (BMI).

The report is written for clinical interpretation. The statistical model
is treated as a robustness analysis rather than as a causal model.

# Abbreviations

**BCM**: body cell mass. **BMI**: body mass index. **BIA**:
bioelectrical impedance analysis. **CI**: confidence interval. **CK**:
creatine kinase. **ENMO**: Euclidean Norm Minus One, an acceleration
metric commonly used for raw accelerometer data. **IIM**: idiopathic
inflammatory myopathy. **LMM**: linear mixed model. **M0, M3, M6, M18,
M30, M48**: study visits at baseline and months 3, 6, 18, 30, and 48.
**PhA**: phase angle. **SE**: standard error. **VIF**: variance
inflation factor.

# Data Overview

The source dataset contains `r nrow(input_data)` observations from
`r dplyr::n_distinct(input_data$filename_2)` patients. The model-eligible
dataset, after deriving between-patient BCM and BMI and removing records
with missing model variables, contains `r nrow(eligible_data)`
observations from `r dplyr::n_distinct(eligible_data$filename_2)`
patients.

```{r data_overview}
tibble::tibble(
  metric = c(
    "Rows in source data",
    "Unique patients in source data",
    "Rows eligible for model",
    "Unique patients eligible for model"
  ),
  value = c(
    nrow(input_data),
    dplyr::n_distinct(input_data$filename_2),
    nrow(eligible_data),
    dplyr::n_distinct(eligible_data$filename_2)
  )
) |>
  gt::gt() |>
  gt::tab_header(title = "Data availability")
```

```{r exam_distribution}
input_data |>
  count(exam, name = "n_source") |>
  left_join(
    eligible_data |> count(exam, name = "n_model_eligible"),
    by = "exam"
  ) |>
  mutate(n_model_eligible = replace_na(n_model_eligible, 0L)) |>
  gt::gt() |>
  gt::tab_header(title = "Visit distribution")
```

```{r descriptive_table}
input_data |>
  summarise(
    age_median = median(age, na.rm = TRUE),
    age_iqr = IQR(age, na.rm = TRUE),
    bmi_median = median(bmi, na.rm = TRUE),
    bmi_iqr = IQR(bmi, na.rm = TRUE),
    bcm_median = median(bcm, na.rm = TRUE),
    bcm_iqr = IQR(bcm, na.rm = TRUE),
    phase_angle_median = median(phase_angle, na.rm = TRUE),
    phase_angle_iqr = IQR(phase_angle, na.rm = TRUE),
    ig_gradient_enmo_median = median(ig_gradient_enmo, na.rm = TRUE),
    ig_gradient_enmo_iqr = IQR(ig_gradient_enmo, na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  gt::gt() |>
  gt::fmt_number(columns = value, decimals = 2) |>
  gt::tab_header(title = "Continuous-variable summary in source data")
```

The raw dataset is longitudinal and unbalanced. Most records are at M0,
M3, and M6; late follow-up is sparse, especially M48. This matters for
interpretation: the M48 coefficient is directionally strong, but it is
estimated from fewer available observations than the earlier visits.

# Methods

The fitted full-sample model was:

```{r model_formula}
if (file.exists(file_full_model)) {
  full_model <- readRDS(file_full_model)
  cat(deparse(stats::formula(full_model)))
}
```

The model uses `ig_gradient_enmo` as the dependent variable. The fixed
effects are visit, age, sex, between-patient BCM, and between-patient
BMI. A random intercept is included for `boot_id`, which represents the
patient or bootstrap cluster. Models were fitted by maximum likelihood
with `lme4::lmer` (`REML = FALSE`), using the `bobyqa` optimizer.

For bootstrap inference, 5000 bootstrap samples were generated. Each
cycle sampled 50 patients with replacement. Within each bootstrap cycle,
the preprocessing was repeated, including derivation of between-patient
components for BCM and BMI using group-mean decomposition. Both raw and
standardized models were fitted. In the standardized model, the response
and continuous predictors were z-standardized within the bootstrap
dataset; therefore, standardized coefficients are comparable by effect
size.

The bootstrap summary reports the mean and median coefficient across
successful bootstrap fits, bootstrap SE, percentile 95% CI, two-sided
bootstrap p value, and the proportion of bootstrap coefficients above or
below zero.

# Model Stability

```{r bootstrap_status}
status_tbl |>
  gt::gt() |>
  gt::tab_header(title = "Bootstrap convergence status") |>
  gt::fmt_number(columns = where(is.numeric), decimals = 0)
```

All 5000 bootstrap cycles converged for both raw and standardized
models. No singular fits were detected. The median bootstrap model used
171 observations and 171 bootstrap clusters. This is a technically
important result: the main findings are not driven by frequent model
failure or unstable random-effect estimation.

# Full-Sample Model

```{r full_sample_parameters}
full_params |>
  select(
    Parameter = parameter_label,
    Coefficient,
    SE,
    `95% CI` = ci,
    t,
    df_error,
    p = p_fmt
  ) |>
  gt::gt() |>
  gt::fmt_number(columns = c(Coefficient, SE, t), decimals = 3) |>
  gt::fmt_number(columns = df_error, decimals = 0) |>
  gt::tab_header(title = "Full-sample fixed effects: raw model")
```

The full-sample model shows a consistent increase in `ig_gradient_enmo`
at all post-baseline visits. Compared with M0, coefficients are positive
at M3, M6, M18, M30, and M48. Age and BMI show negative associations,
whereas male sex and BCM do not show clear independent effects in the
full-sample model.

```{r collinearity}
full_collin |>
  gt::gt() |>
  gt::fmt_number(columns = where(is.numeric), decimals = 2) |>
  gt::tab_header(title = "Collinearity diagnostics for the full-sample model")
```

VIF values are below conventional high-collinearity thresholds. BCM and
sex have the highest VIFs, but neither reaches a level that would by
itself invalidate the model. The covariate estimates should still be
read as mutually adjusted associations, not as isolated biological
effects.

# Bootstrap Results

```{r bootstrap_raw}
summary_tbl |>
  filter(model_type == "raw") |>
  select(
    Parameter = parameter_label,
    `Median coefficient` = estimate,
    `Bootstrap SE` = boot_se_fmt,
    `95% percentile CI` = ci,
    `Bootstrap p` = p_fmt,
    `Pr(coef > 0)` = prop_positive,
    `Pr(coef < 0)` = prop_negative,
    Direction = direction
  ) |>
  gt::gt() |>
  gt::fmt_number(columns = c(`Pr(coef > 0)`, `Pr(coef < 0)`), decimals = 3) |>
  gt::tab_header(title = "Bootstrap summary: raw coefficients")
```

The raw bootstrap model confirms that post-baseline visits have positive
coefficients. The 95% percentile CI excludes zero for M3, M6, M18, M30,
and M48. The estimated raw increase is approximately 0.062 at M3, 0.070
at M6, 0.105 at M18, 0.102 at M30, and 0.162 at M48, all relative to
M0. Because `ig_gradient_enmo` is on a transformed accelerometry scale,
the most useful clinical reading is the direction and robustness of the
change rather than a direct bedside unit conversion.

Age has a small but robust negative coefficient. BMI also has a robust
negative between-patient association. BCM is mostly positive but its
percentile CI crosses zero; therefore, the data do not support a
definitive independent BCM effect in this model. Male sex is not
estimated with sufficient precision to support a sex-specific
conclusion.

```{r bootstrap_standardized}
summary_tbl |>
  filter(model_type == "standardized") |>
  select(
    Parameter = parameter_label,
    `Median coefficient` = estimate,
    `Bootstrap SE` = boot_se_fmt,
    `95% percentile CI` = ci,
    `Bootstrap p` = p_fmt,
    `Pr(coef > 0)` = prop_positive,
    `Pr(coef < 0)` = prop_negative,
    Direction = direction
  ) |>
  gt::gt() |>
  gt::fmt_number(columns = c(`Pr(coef > 0)`, `Pr(coef < 0)`), decimals = 3) |>
  gt::tab_header(title = "Bootstrap summary: standardized coefficients")
```

The standardized model makes the relative effect sizes easier to
compare. The strongest standardized visit coefficient is M48, followed
by M18 and M30, but this must be tempered by the sparse M48 data. Age
and BMI have similar standardized negative associations
(approximately -0.22 and -0.24 SD, respectively). BCM is positive
(approximately +0.21 SD) but statistically uncertain because its
bootstrap interval includes zero.

# Figures

```{r figures_raw, fig.cap='Raw-model bootstrap coefficient distributions and percentile confidence intervals.'}
knitr::include_graphics(c(
  file.path(path_boot, "bootstrap_density_raw.png"),
  file.path(path_boot, "bootstrap_forest_raw.png")
))
```

```{r figures_standardized, fig.cap='Standardized-model bootstrap coefficient distributions and percentile confidence intervals.'}
knitr::include_graphics(c(
  file.path(path_boot, "bootstrap_density_standardized.png"),
  file.path(path_boot, "bootstrap_forest_standardized.png")
))
```

The density plots show the empirical bootstrap distribution for each
fixed-effect coefficient. A distribution concentrated on one side of
zero supports directional robustness. The forest plots summarize the
median bootstrap coefficient and the 95% percentile CI. The visual
pattern is consistent with the tables: visit effects are consistently
positive, age and BMI are consistently negative, while BCM and sex are
less certain.

# Table and File Dictionary

The folder contains both human-readable and machine-readable versions of
the results:

- `bootstrap_lmer_report.xlsx`: Excel workbook with seven sheets. The
  most important sheets are `full_sample_parameters_raw`,
  `full_sample_collinearity_raw`, `bootstrap_summary`,
  `bootstrap_status`, `bootstrap_all_parameters`, and `failed_boots`.
- `bootstrap_summary.csv` / `bootstrap_summary.rds`: compact bootstrap
  summary by model type and fixed-effect parameter.
- `bootstrap_all_parameters.csv` / `bootstrap_all_parameters.rds`:
  coefficient-level output from every bootstrap model. This table is the
  audit trail for the bootstrap distributions.
- `bootstrap_status.csv` / `bootstrap_status.rds`: convergence and
  singularity summary for raw and standardized models.
- `full_sample_model_raw.rds`: fitted `lmerMod` object for the
  full-sample raw model.
- `bootstrap_density_raw.png` and
  `bootstrap_density_standardized.png`: density plots of bootstrap
  coefficient distributions.
- `bootstrap_forest_raw.png` and
  `bootstrap_forest_standardized.png`: forest plots of median bootstrap
  coefficients with percentile CI.

Column definitions:

- `model_type`: raw or standardized model. In the standardized model,
  continuous predictors and the response are z-standardized.
- `Parameter`: model coefficient. `examM3`, `examM6`, `examM18`,
  `examM30`, and `examM48` are visit contrasts relative to M0.
  `sexMale` is male sex relative to female. `bcm_between` and
  `bmi_between` are between-patient components of BCM and BMI.
- `n_success`: number of successful bootstrap fits contributing to the
  parameter summary.
- `estimate_mean`, `estimate_median`: mean and median bootstrap
  coefficient.
- `boot_se`: standard deviation of the bootstrap coefficient
  distribution.
- `ci_low_percentile`, `ci_high_percentile`: 2.5th and 97.5th
  percentiles of the bootstrap coefficient distribution.
- `p_boot_two_sided`: empirical two-sided bootstrap p value calculated
  from the proportion of coefficients on either side of zero.
- `prop_positive`, `prop_negative`: proportion of bootstrap coefficients
  greater than or less than zero.
- `Coefficient`, `SE`, `CI_low`, `CI_high`, `t`, `df_error`, `p`: fixed
  effect estimate and model-based uncertainty from a single fitted
  model.
- `VIF`: variance inflation factor. Higher values indicate stronger
  collinearity with other predictors.
- `Tolerance`: reciprocal information to VIF; lower values indicate
  stronger collinearity.
- `boot_id`: bootstrap replicate number in the coefficient table; in
  the fitted model formula, `boot_id` also identifies the patient or
  bootstrap cluster for the random intercept.
- `converged`, `singular`, `error_message`: model fit diagnostics.
- `n_obs`, `n_groups`: number of model observations and random-effect
  groups used in a fitted bootstrap model.

# Clinical Interpretation

From a clinical perspective, the most reproducible signal is the
increase in accelerometer-derived `ig_gradient_enmo` after baseline.
This supports the interpretation that longitudinal accelerometry can
capture change in physical behavior or functional status in IIM. The
finding is biologically plausible because accelerometer-derived physical
activity has been proposed as a useful outcome domain in IIM, where
standard clinical measures may not fully capture day-to-day functional
capacity. A review of accelerometer-derived physical activity in IIM
concluded that the field is still developing, but that accelerometry has
promise for disease-activity and intervention assessment
([Oldroyd et al., 2019](https://pmc.ncbi.nlm.nih.gov/articles/PMC6805320/)).
An adult IIM study also described physical activity monitoring as a
promising outcome measure
([Bachasson et al., 2017](https://pubmed.ncbi.nlm.nih.gov/28566549/)).

The negative age effect indicates that older patients tend to have lower
accelerometry signal after adjustment for visit, sex, BCM, and BMI. This
is clinically expected and should be treated as an important confounder
in future models of activity and muscle status.

The negative BMI effect is also clinically meaningful. Higher BMI can
reduce ambulatory activity, alter wrist-accelerometer signal, and partly
mask muscle-quality or disease-activity relationships. Because the model
uses the between-patient BMI component, this result should be read as:
patients with higher average BMI across follow-up tend to have lower
`ig_gradient_enmo`, independent of visit and the other covariates.

BCM is directionally positive but uncertain. This fits the clinical
expectation that greater cellular/muscle mass may relate to better
activity, but the current dataset does not provide strong independent
evidence once BMI, age, sex, and visit are included. In a manuscript,
this should be presented as supportive but not confirmatory.

Phase angle is not the dependent variable in this bootstrap model, but
it is central to the broader biological context. BIA-derived phase angle
is commonly interpreted as a marker related to cell membrane integrity,
fluid distribution, nutritional status, and muscle quality. Reviews and
clinical studies support its relevance for muscle quality and sarcopenia
assessment
([Norman et al., 2012](https://pubmed.ncbi.nlm.nih.gov/22698802/);
[Akamatsu et al., 2022](https://pubmed.ncbi.nlm.nih.gov/34845859/);
[Di Vincenzo et al., 2021](https://pubmed.ncbi.nlm.nih.gov/33183880/)).
In IIM, where inflammation, muscle damage, glucocorticoid exposure,
fatigue, and deconditioning can coexist, phase angle and accelerometry
should be interpreted as complementary functional-bioelectrical readouts
rather than interchangeable measures.

# Manuscript-Oriented Summary

In a longitudinal IIM dataset of 64 patients and 219 observations,
bootstrap mixed-effects modelling showed robust positive visit effects
for `ig_gradient_enmo` after baseline. The analysis used 5000 bootstrap
cycles with patient-level resampling and repeated preprocessing within
each cycle. All raw and standardized bootstrap models converged without
singular fits. Compared with M0, visit coefficients were consistently
positive from M3 through M48. Age and between-patient BMI were
independently and robustly negatively associated with
`ig_gradient_enmo`. Between-patient BCM showed a positive but
statistically uncertain association, and male sex was not clearly
associated with the outcome.

These results support the use of accelerometer-derived activity
measures as a longitudinal functional readout in IIM. The findings also
indicate that age and BMI must be retained as clinically relevant
covariates in models linking activity, body composition, and phase
angle. The sparse late follow-up, particularly M48, should be explicitly
acknowledged as a limitation.

# References

- Akamatsu Y, Kusakabe T, Arai H, et al. Phase angle from
  bioelectrical impedance analysis is a useful indicator of muscle
  quality. *Journal of Cachexia, Sarcopenia and Muscle*. 2022;13:180-189.
  doi: [10.1002/jcsm.12860](https://pubmed.ncbi.nlm.nih.gov/34845859/).
- Bachasson D, Landon-Cardinal O, Benveniste O, Hogrel JY, Allenbach Y.
  Physical activity monitoring: A promising outcome measure in
  idiopathic inflammatory myopathies. *Neurology*. 2017;89:101-103.
  doi: [10.1212/WNL.0000000000004061](https://pubmed.ncbi.nlm.nih.gov/28566549/).
- Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
  models using lme4. *Journal of Statistical Software*. 2015;67:1-48.
  doi: [10.18637/jss.v067.i01](https://www.jstatsoft.org/v67/i01/).
- Di Vincenzo O, Marra M, Di Gregorio A, Pasanisi F, Scalfi L.
  Bioelectrical impedance analysis-derived phase angle in sarcopenia: A
  systematic review. *Clinical Nutrition*. 2021;40:3052-3061.
  doi: [10.1016/j.clnu.2020.10.048](https://pubmed.ncbi.nlm.nih.gov/33183880/).
- Efron B. Bootstrap methods: Another look at the jackknife. *The
  Annals of Statistics*. 1979;7:1-26. doi:
  [10.1214/aos/1176344552](https://doi.org/10.1214/aos/1176344552).
- Norman K, Stobäus N, Pirlich M, Bosy-Westphal A. Bioelectrical phase
  angle and impedance vector analysis: clinical relevance and
  applicability of impedance parameters. *Clinical Nutrition*.
  2012;31:854-861. doi:
  [10.1016/j.clnu.2012.05.008](https://pubmed.ncbi.nlm.nih.gov/22698802/).
- Oldroyd A, Little MA, Dixon W, Chinoy H. A review of
  accelerometer-derived physical activity in the idiopathic inflammatory
  myopathies. *BMC Rheumatology*. 2019;3:41. doi:
  [10.1186/s41927-019-0088-1](https://pmc.ncbi.nlm.nih.gov/articles/PMC6805320/).

# Session Info

```{r session_info}
sessioninfo::session_info()
```
